
---
title: "PA1_template"
author: "Karthik Chawala"
date: "August 31, 2017"
output:
  md_document: default
  html_document: default
---


# Reproducable Research Course Project 1



## Load the necessary packages and set default of echo to true

```{r}
library(knitr)
opts_chunk$set(echo = TRUE)
library(dplyr)
library(lubridate)
library(ggplot2)
```


## Loading and preprocessing the data

### Read the Data
```{r}
if(!file.exists('activity.csv'))
  {
  unzip('activity.zip')
  }
data <- read.csv("activity.csv", header = TRUE, sep = ',', colClasses = c("numeric", "character", "integer"))
```

### Process the Data
```{r}
data$date <- ymd(data$date)
```



## What is mean total number of steps taken per day?

### 1. Calculate the total number of steps taken per day
```{r}
steps <- data %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print
```

### 2. Make a histogram of the total number of steps taken each day
```{r}
ggplot(steps, aes(x = steps)) +
  geom_histogram(fill = "black", binwidth = 1000) +
  labs(title = "Total Steps per day", x = "Steps per day", y = "Frequency")
```

### 3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
mean_steps <- mean(steps$steps, na.rm = TRUE)
median_steps <- median(steps$steps, na.rm = TRUE)

mean_steps
median_steps
```




## What is the average daily activity pattern?

### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
interval <- data %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(steps = mean(steps))

ggplot(interval, aes(x=interval, y=steps)) +
  geom_line(color = "black")
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
highestSteps <- interval[which.max(interval$steps),]
print(highestSteps)
```




## Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
nbrMissingValues <- sum(is.na(data$steps))
print(nbrMissingValues)
```


### 2. Devise a strategy for filling in all of the missing values in the dataset. 

### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
data_full <- data
nas <- is.na(data_full$steps)
avg_interval <- tapply(data_full$steps, data_full$interval, mean, na.rm=TRUE, simplify=TRUE)
data_full$steps[nas] <- avg_interval[as.character(data_full$interval[nas])]
```

### 4. Make a histogram of the total number of steps taken each day
```{r}
steps_full <- data_full %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print

ggplot(steps_full, aes(x = steps)) +
  geom_histogram(fill = "black", binwidth = 1000) +
  labs(title = "Steps per day, including missing values", x = "Steps per day", y = "Frequency")
```

### Calculate and report the mean and median total number of steps taken per day
```{r}
mean_steps_full <- mean(steps_full$steps, na.rm = TRUE)
median_steps_full <- median(steps_full$steps, na.rm = TRUE)

mean_steps_full
median_steps_full

```



## Are there differences in activity patterns between weekdays and weekends?

### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
data_full <- mutate(data_full, weektype = ifelse(weekdays(data_full$date) == "Saturday" | weekdays(data_full$date) == "Sunday", "weekend", "weekday"))
data_full$weektype <- as.factor(data_full$weektype)
head(data_full)
```

### Make a panel plot containing a time series plot
```{r}
interval_full <- data_full %>%
  group_by(interval, weektype) %>%
  summarise(steps = mean(steps))
s <- ggplot(interval_full, aes(x=interval, y=steps, color = weektype)) +
  geom_line() +
  facet_wrap(~weektype, ncol = 1, nrow=2)
print(s)
```
